---
title: 'Bios 6301: Assignment 2'
output: pdf_document
author: Valerie Welty
---

*(informally) Due Tuesday, 20 September, 1:00 PM*

50 points total.

This assignment won't be submitted until we've covered Rmarkdown.
Create R chunks for each question and insert your R code appropriately.
Check your output by using the `Knit PDF` button in RStudio.

1. **Working with data** In the `datasets` folder on the course GitHub repo, you will find a file called `cancer.csv`, which is a dataset in comma-separated values (csv) format. This is a large cancer incidence dataset that summarizes the incidence of different cancers for various subgroups. (18 points)

    1. Load the data set into R and make it a data frame called `cancer.df`. (2 points)
    
    ```{r, tidy=TRUE}
    cancer.df <- as.data.frame(read.csv("cancer.csv"))
    ```

    2. Determine the number of rows and columns in the data frame. (2)
    
    ```{r, tidy=TRUE}
    nrow(cancer.df)
    ncol(cancer.df)
    ```

    3. Extract the names of the columns in `cancer.df`. (2)
    ```{r, tidy=TRUE}
    (head <- names(cancer.df))
    ```

    4. Report the value of the 3000th row in column 6. (2)
    ```{r, tidy=TRUE}
    cancer.df[3000,6]
    ```

    5. Report the contents of the 172nd row. (2)
    ```{r, tidy=TRUE}
    cancer.df[172,]
    ```

    6. Create a new column that is the incidence *rate* (per 100,000) for each row.(3)
    ```{r, tidy=TRUE}
    cancer.df[,9] <- cancer.df[,7]/100000
    names(cancer.df)[9] = 'rate'
    head(cancer.df)
    ```

    7. How many subgroups (rows) have a zero incidence rate? (2)
    ```{r, tidy=TRUE}
    head(table(cancer.df[,9]))[1]
    # 23,191 subgroups have a zero incidence rate.
    ```

    8. Find the subgroup with the highest incidence rate.(3)
    ```{r, tidy=TRUE}
    (max <- cancer.df[which.max(cancer.df[,9]),])
    # The highest incidence rate in the data set is 0.18744, occuring at subgroup 21387.  
    ```

2. **Data types** (10 points)

    1. Create the following vector: `x <- c("5","12","7")`. Which of the following commands will produce an error message? For each command, Either explain why they should be errors, or explain the non-erroneous result. (4 points)
    
    max(x)
    
    sort(x)
    
    sum(x)
    
    ```{r, tidy=TRUE}
    x <- c("5","12","7")
    mode(x)
    max(x)
    # max(x) will report the last value in the vector for character type. 
    sort(x)
    # sort(x) will sort the values alphabetically, hence why "12" occurs first (the 1 in front)
    
    # sum(x)
    
    # sum(x) will report an error because the mode of the variable x is character, and sum(x) can only operate on numeric/integer types.
    ```

            
    2. For the next two commands, either explain their results, or why they should produce errors. (3 points)

            y <- c("5",7,12)
            
            y[2] + y[3]
            
    ```{r, tidy=TRUE}
    y <- c("5",7,12)
    mode(y)
      # y[2] + y[3]
    
    # this operation should produce an error because, while 7 and 12 were inputted as numeric values, there was a character value in the vector as well. Character is the least flexible value, so because there was a character value in the vector the remaining values will be forced into character values as well. Thus, y[2] and y[3] are not numeric values and thus cannot be operated on mathematically. 
    ```

    3. For the next two commands, either explain their results, or why they should produce errors. (3 points)

            z <- data.frame(z1="5",z2=7,z3=12)
            
            z[1,2] + z[1,3]
            
    ```{r, tidy=TRUE}
    z <- data.frame(z1="5",z2=7,z3=12)
            z[1,2] + z[1,3]
    # The command will take the value in the 1st row and 2nd column of z (=7) and add it to the value in the first row and third column of z (=12) which results in 19. 
    ```

3. **Data structures** Give R expressions that return the following matrices and vectors (*i.e.* do not construct them manually). (3 points each, 12 total)

    1. $(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)$
    ```{r, tidy=TRUE}
    x <- c(seq(1:8), rev(1:7))
    x
    ```
    
    2. $(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)$
    ```{r, tidy=TRUE}
    x <- rep(1:5, times=1:5)
    x
    ```

    3. $\begin{pmatrix}
      0 & 1 & 1  \\
      1 & 0 & 1  \\
      1 & 1 & 0  \\
    \end{pmatrix}$
    
    ```{r, tidy=TRUE}
    x <- matrix(data=1, nrow=3, ncol=3)
    diag(x) = 0
    x
    ```

    4. $\begin{pmatrix}
      1 & 2 & 3 & 4 \\
      1 & 4 & 9 & 16 \\
      1 & 8 & 27 & 64  \\
      1 & 16 & 81 & 256 \\
      1 & 32 & 243 & 1024  \\
    \end{pmatrix}$
    
    ```{r, tidy=TRUE}
    x <- matrix(data=NA, nrow=5, ncol=4)
    for (i in 1:5) {for (j in 1:4) {x[i,j] = j^i}}
    x
    ```

4. **Basic programming** (10 points)

    1. Let $h(x,n)=1+x+x^2+\ldots+x^n = \sum_{i=0}^n x^i$. Write an R program to calculate $h(x,n)$ using a `for` loop. (5 points)
    ```{r, tidy=TRUE}
    h <- function(x,n) {
      sum <- 0
      for (i in 0:n) {sum <- sum + (x^i)}
      return(sum)
    }
    
    h(2,3)
    h(4,5)
    h(6,3)
    ```

    1. If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Write an R program to perform the following calculations. (5 points)
    
    ```{r, tidy=TRUE}
    euler <- function(x.1, x.2, n) {
    x <- vector(mode="integer", length=0)
    sum <- 0
    for (i in 1:n-1) {
      if ((i %% x.1 == 0) | (i %% x.2 == 0)) {
        x <- c(x,i)
        sum <- sum + i
      }
    }
    x
    return(sum)
    }
    ```

        1. Find the sum of all the multiples of 3 or 5 below 1,000. (3, [euler1])
        
    ```{r, tidy=TRUE}
    euler(3,5,1000)
    ```

        1. Find the sum of all the multiples of 4 or 7 below 1,000,000. (2)
        
    ```{r, tidy=TRUE,cache=TRUE}
    euler(4,7,1000000)
    ```

    1. Each new term in the Fibonacci sequence is generated by adding the previous two terms. By starting with 1 and 2, the first 10 terms will be $(1, 2, 3, 5, 8, 13, 21, 34, 55, 89)$. Write an R program to calculate the sum of the first 15 even-valued terms. (5 bonus points, [euler2])

    ```{r, tidy=TRUE}
    sumeven <- function(n) {
    x <- c(1,2)
    even <- vector(mode="integer", length=0L)
      for (i in 3:100) {
          x[i] = x[i-1] + x[i-2]
          if (x[i] %% 2 == 0) {even = c(even, x[i])}
          if (length(even)==n) {break}
        }
    print(x)
    print(even)
    print(paste0('sum of even fibonacci values up to ', n, ' = '))
    return(sum(even))
    }
    sumeven(15)
    ```



Some problems taken or inspired by projecteuler.

[euler1]:https://projecteuler.net/problem=1
[euler2]: https://projecteuler.net/problem=2